Smoothing Algorithms for Variable Rate Models

### 2012-03-05 draft1_VRsmoothing.tex
merged filtering and rb smoothing

\usepackage{cite}
\usepackage[pdftex]{graphicx}
\usepackage[cmex10]{amsmath}
\usepackage{amsfonts}
\usepackage{algorithmic}
\usepackage{array}
\usepackage{mdwmath}
\usepackage{mdwtab}
\usepackage{eqparbox}
\usepackage[caption=false,font=footnotesize]{subfig}
\usepackage{fixltx2e}
\usepackage{url}
\usepackage{color}

@@

{\meta Add a bit about structure of paper.}

@@

 which has values $\{ x_0, x_1 \dots x_K \}$ at the changepoints. 
 Conditional upon a realisation of this process, $x(t)$ may be calculated deterministically.
   The cardinality of these sets (i.e. the number of changepoints occurring in the appropriate range) is denoted $K_n$, $K_n^+$ and $K_{n \setminus n-1}$ respectively.
%Note that $\theta_n$ not only conveys information about where changepoints occur, but also where they do not. i.e. it is implicit that $\tau_{K_n+1} > t_n$.

The changepoint sequence is assumed to be a Markov process.

\begin{IEEEeqnarray}{rCl}
 \{\tau_k, u_k\} & \sim & p(u_k|\tau_k, \tau_{k-1}, u_{k-1}) p(\tau_k|\tau_{k-1}, u_{k-1}) \label{eq:cp_model}
\end{IEEEeqnarray}

In the manner of \cite{Whiteley2011}, a survivor function is defined as the probability that no new changepoint occurs before a given time.

\begin{IEEEeqnarray}{rCl}
 S(\tau_k, u_k, t) &=& P(\tau_{k+1}>t|\tau_k, u_k) \nonumber \\
              &=& 1 - \int_{\tau_k}^{t} p(\xi|\tau_{k}, u_k) d\xi
\end{IEEEeqnarray}

@@

\begin{IEEEeqnarray}{rCl}
p(\theta_n) & = & S(\tau_{K_n},t_n) p(u_0) \prod_{k=1}^{K_n} p(\tau_k, u_k| \tau_{k-1}, u_{k-1}) \label{eq:cp_sequence_prior}  .
\end{IEEEeqnarray}

Next we consider the state dynamics. A counting variable is introduced to keep track of the most recent changepoint to have occurred.

\begin{IEEEeqnarray}{rCl}
 \nu_t = \max(k : \tau_k<t)
\end{IEEEeqnarray}

Conditional upon the sequence of past changepoints, $\theta_n$, the state dynamics are given by a deterministic differential equation.

\begin{IEEEeqnarray}{rCl}
 dx(t) & = & \mathrm{f}(x(t), \tau_{\nu_t}, u_{\nu_t})
\end{IEEEeqnarray}

Assuming an analytic solution exists, a state transition function may be calculated.

\begin{IEEEeqnarray}{rCll}
 x(t) & = & f(x_{\nu_t}, v_{\nu_t}, \tau_{\nu_t}, t) &, \tau_{\nu_t} < t <= \tau_{\nu_t+1}    \label{eq:disc_time_state_diff_eq}
\end{IEEEeqnarray}

By choosing $t = \tau_{\nu_t+1}$, this equation specifies the state at the next changepoint time. Similarly, by choosing $t=t_n$, the state at the observation times is specified. These points will be denoted $\hat{x}_n$. Note than these points can be calculated deterministically given the initial state and past changepoint sequence.

\begin{IEEEeqnarray}{rCll}
 \hat{x}_n & = & f(x_0, \theta_n, t)   \label{eq:obs_time_state}
\end{IEEEeqnarray}

It is assumed for notational simplicity in the following sections that the initial state, $x_0$, is known a priori. This condition is easily relaxed.

In a target tracking example, $\tau_{1:K}$ might specify the times at which manoeuvres start, and $u_{1:K}$ the accelerations or other motion parameters associated with each manoeuvre. $x(t)$ would be the kinematic state of the target.

We consider a general model from time $0$ to $T$, between which observations, $\{y_1 \dots y_N\}$, are made at times $\{t_1 \dots t_N = T\}$. The linear state at these times is written as $\{x_1 \dots x_N\}$. During this period, an unknown number of changepoints, $K$, occur at times $\{ \tau_1 \dots \tau_K \}$, each with associated changepoint parameters, $\{ u_1 \dots u_K \}$. These are the marks of the MPP. Discrete sets containing multiple values over time will be written as, e.g. $y_{1:n} = \{y_1 \dots y_n\}$.

The objective for inference will be to estimate the sequence of changepoints. This will be denoted as $\theta = \{\tau_{1:K}, u_{1:K}\}$. In addition, at a particular time $t_n$, the sequence will be divided into past $\theta_n = \{\tau_{j}, u_{j} \forall j : 0 \leq \tau_j < t_n \}$, and future $\theta_n^+ = \{\tau_{j}, u_{j} \forall j : t_n \leq \tau_j < T \}$. It will also be useful to define a variable for the changepoints which occur in the interval $[t_{n-1},t_n)$, $\theta_{n \setminus n-1} = \{\tau_{j}, u_{j} \forall j : t_{n-1} \leq \tau_j < t_n \}$. The cardinality of these sets (i.e. the number of changepoints occurring in the appropriate range) is denoted $K_n$, $K_n^+$ and $K_{n \setminus n-1}$ respectively.

Note that $\theta_n$ not only conveys information about where changepoints occur, but also where they do not. i.e. it is implicit that $\tau_{K_n+1} > t_n$.%; furthermore that $\theta_n$ conveys less information than the two sets $\theta_{n-1}$ and $\theta_{n \setminus n-1}$ separately. The latter also impose the condition that $\tau_{K_{n-1}} < t_{n-1} < \tau_{K_{n-1}+1}$.

Assuming a Markovian changepoint sequence, a conditionally linear-Gaussian variable rate model can now be expressed by the following system equations:
%
\begin{IEEEeqnarray}{rCl}
 \{\tau_k, u_k\} & \sim & p(u_k|\tau_k, \tau_{k-1}, u_{k-1}) p(\tau_k|\tau_{k-1}) \label{eq:cp_model} \\
 x_n & = & A_n(\theta_{n})x_{n-1} + w_n \\
 y_n & = & C_n(\theta_{n})x_n + v_n  .
\end{IEEEeqnarray}

The random variables $w_n$ and $v_n$ have a zero-mean Gaussian distribution with covariance matrices $Q_n$ and $R_n$ respectively. The changepoint density will be constructed such that $P(\tau_k < \tau_{k-1}) = 0$. The factorisation of (\ref{eq:cp_model}) may be relaxed, but is assumed throughout as it simplifies the subsequent derivations.

It will be useful to define the survivor function in the same manner as \cite{Whiteley2011}.
%
\begin{IEEEeqnarray}{rCl}
 S(\tau_k, t) &=& P(\tau_{k+1}>t|\tau_k) \nonumber \\
              &=& 1 - \int_{\tau_k}^{t} p(\xi|\tau_{k}) d\xi
\end{IEEEeqnarray}

Using the convention that $\tau_0 = 0$ and $u_0$ is known a priori, the prior distribution over the changepoint sequence, $\theta_n$, may now be expressed as,
%
\begin{IEEEeqnarray}{rCl}
p(\theta_n) & = & S(\tau_{K_n},t_n) \prod_{k=1}^{K_n} p(\tau_k, u_k| \tau_{k-1}, u_{k-1}) \label{eq:cp_sequence_prior}  .
\end{IEEEeqnarray}

See \cite{Jacobsen2006} for a rigorous derivation of MPP probability measures.

The tasks of filtering and smoothing may now be considered as estimation of the posterior distributions of $\theta$ and $x_{1:N}$.

\end{document}

Target tracking algorithms are commonly based upon fixed rate models (see, e.g. \cite{Li2003} for a thorough survey), in which the target kinematics (position, velocity, etc.) are estimated at a set of fixed times at which observations (e.g. radar measurements) are made. In \cite{Godsill2007a,Godsill2007}, variable rate models were introduced for tracking, in which the state trajectory is divided up by a set of changepoints between which the motion follows a deterministic path governed by motion parameters (accelerations, etc.) which are fixed for that division. The inference problem is hence reduced to estimation of the changepoint times and associated motion parameters. As noted in \cite{Whiteley2011}, these comprise a marked point process (MPP), the mathematical properties of which are thoroughly set out in \cite{Jacobsen2006}.

We introduce piecewise-deterministic variable rate models in section~\ref{sec:pd_vr_models}, and review the VRPF in section~\ref{sec:pd_vrpf}. The new variable rate particle smoother (VRPS) algorithm is described in section~\ref{sec:pd_vrps}. In section~\ref{sec:tracking_models}, we adapt the intrinsic-coordinate models of \cite{Godsill2007a,Godsill2007} for use with the new smoothing algorithm. Simulations are presented in section~\ref{sec:simuations}.

The posterior distribution of the changepoint sequence is analytically intractable, but may be estimated using a particle filter. When the conditional state dynamics are linear-Gaussian, the efficiency of this filter may be greatly improved by using the method of Rao-Blackwellisation (see, e.g. \cite{Casella1996,Doucet2000}). Rather than targeting the entire posterior distribution over changepoints and states, the particle filter is used to estimate only the nonlinear changepoint sequence, after which Kalman filtering and smoothing may be used to infer the linear states. This is the Rao-Blackwellised variable rate particle filter (RBVRPF) of \cite{Godsill2007a,Christensen2012}.

In this paper, a new algorithm is described for use with conditionally linear-Gaussian variable rate models for estimation of the smoothing distribution, i.e. the distribution over the sequence of changepoints and linear states given all the observations. This uses a similar derivation to that of the fixed rate Rao-Blackwellised particle smoother of \cite{Sarkka2012}. The new algorithm is called the Rao-Blackwellised variable rate particle smoother (RBVRPS).

We review the structure of conditionally linear-Gaussian variable rate models in section~\ref{sec:rbvr_models} and revise the RBVRPF in section~\ref{sec:rbvrpf}. The new smoothing algorithm is presented in section~\ref{sec:rbvrps} with supporting simulations in section~\ref{sec:simulations}.

>>>>

\usepackage{cite}
\usepackage[pdftex]{graphicx}
\usepackage[cmex10]{amsmath}
\usepackage{amsfonts}
\usepackage{algorithmic}
\usepackage{array}
\usepackage{mdwmath}
\usepackage{mdwtab}
%\usepackage{eqparbox}
\usepackage[caption=false,font=footnotesize]{subfig}
\usepackage{fixltx2e}
\usepackage{url}
\usepackage{color}

@@

We introduce the general structure of variable rate models in section~\ref{sec:vr_models} and discuss two particular types, those with conditionally linear-Gaussian and conditionally deterministic dynamics. The VRPF is reviewed in section~\ref{sec:vrpf}, and new variable rate smoothing algorithms are described in section~\ref{sec:vrps}. In section~\ref{sec:simuations}, particular examples of variable rate models are presented and their performance demonstrated in a series of simulations.

@@

For notational simplicity, the following counting variables are introduced to keep track of the most recent changepoint to have occurred,
%
\begin{IEEEeqnarray}{rCl}
 K(t)  & = & \max(k : \tau_k<t) \\
 K_n   & = & K(t_n)     .
\end{IEEEeqnarray}

The changepoint sequence is assumed to be a Markov process.
%
\begin{IEEEeqnarray}{rCl}
 \{\tau_k, u_k\} & \sim & p(u_k|\tau_k, \tau_{k-1}, u_{k-1}) p(\tau_k|\tau_{k-1}, u_{k-1}) \label{eq:cp_model}
\end{IEEEeqnarray}

The changepoint density will be constructed such that $P(\tau_k < \tau_{k-1}) = 0$.

In the manner of \cite{Whiteley2011}, a survivor function is defined as the probability that no new changepoint occurs before a given time,
%
\begin{IEEEeqnarray}{rCl}
 S(\tau_k, u_k, t) &=& P(\tau_{k+1}>t|\tau_k, u_k) \nonumber \\
              &=& 1 - \int_{\tau_k}^{t} p(\xi|\tau_{k}, u_k) d\xi     .
\end{IEEEeqnarray}

@@

\begin{IEEEeqnarray}{rCl}
p(\theta_n) & = & S(\tau_{K_n},t_n) p(u_0) \prod_{k=1}^{K_n} p(\tau_k, u_k| \tau_{k-1}, u_{k-1}) \label{eq:cp_sequence_prior}
\end{IEEEeqnarray}

\subsection{Conditionally Linear-Gaussian Models}

The first class of variable rate models to be considered is those whose state dynamics are linear-Gaussian conditional on the changepoint sequence. Such a model may be discretised onto the set of observation times in exactly the same manner as a standard the standard HMM.
%
\begin{IEEEeqnarray}{rCl}
 x_n & = & A_n(\theta_{n})x_{n-1} + w_n \\
 y_n & = & C_n(\theta_{n})x_n + v_n       ,
\end{IEEEeqnarray}

where,
%
\begin{IEEEeqnarray}{rCl}
 w_n & \sim & \mathcal{N}(w_n|0, Q_n(\theta_{n})) \\
 v_n & \sim & \mathcal{N}(v_n|0, R_n(\theta_{n}))       .
\end{IEEEeqnarray}

In addition, the prior state distribution should be Gaussian, with known mean and variance.
%
\begin{IEEEeqnarray}{rCl}
 x_0 & \sim & \mathcal{N}(x_0|m_0, P_0)       .
\end{IEEEeqnarray}

If the changepoint sequence is known, or has been estimated, then the state values, $x_n$ may be inferred using optimal Kalman filtering and smoothing recursions. As well as the basic Kalman filter \cite{Kalman1960}, the Rauch-Tung-Striebel (RTS) smoother \cite{Rauch1965} and two-filter smoother \cite{Fraser1969} will prove useful for this step.

Conditionally linear-Gaussian variable rate models were introduced in \cite{Godsill2007a} for a financial inference algorithm. Changepoints correspond to jumps in the value or trend of a security, at which points the process covariance, $Q_n(\theta_n)$, is inflated.

\subsection{Conditionally Deterministic Models}

The second class of variable rate models for consideration is those in which the state is completely specified by the changepoint sequence, with no additional random components. Such a process is commonly referred to as ``piecewise-deterministic'', as the latent state follows a deterministic path between changepoints. In this case, it is not necessary to discretise the state -- it may be kept as a continuous variable.

In general, the state dynamics will be governed by a differential equation which may depend on the entire changepoint sequence. Here we assume that only the most recent changepoint is significant.
%
\begin{IEEEeqnarray}{rCl}
 dx(t) & = & \mathrm{f}(x(t), \tau_{K(t)}, u_{K(t)})     .
\end{IEEEeqnarray}

By introducing a new sequence, $\{ x_0, x_1 \dots x_K \}$, which denotes the value of the state at each changepoint (i.e. $x(\tau_k)$), and assuming that an analytic solution exists, a state transition function may be found,
%
\begin{IEEEeqnarray}{rCll}
 x(t) & = & f(x_{K_n}, v_{K_n}, \tau_{K_n}, t) &, \tau_{K_n} < t <= \tau_{K_{n}+1}    \label{eq:disc_time_state_diff_eq}     .
\end{IEEEeqnarray}

By choosing $t = \tau_{K_{n}+1}$, this equation specifies the state at the next changepoint time. Similarly, by choosing $t=t_n$, the state at the observation times may be calculated. These points will be denoted $\hat{x}_n$. To complete the model, a probabilistic measurement model must be chosen for the observation process, $p(y_n|\hat{x}_n)$.

For convenience, we assume that $x_0$ is known in the following sections. This means that $x(t)$ may be calculated deterministically for all $t$ given $\theta$. However, this condition is easily relaxed by including $x_0$ as a random variable in the posterior distribution.

Target tracking algorithms are commonly based upon fixed rate models (see, e.g. \cite{Li2003} for a thorough survey), in which the target kinematics (position, velocity, etc.) are estimated at a set of fixed times at which observations (e.g. radar measurements) are made. In \cite{Godsill2007a,Godsill2007}, variable rate models were introduced for tracking, in which the state trajectory is divided up by a set of changepoints between which the motion follows a deterministic path governed by motion parameters (accelerations, etc.) which are fixed for that division.

\section{The Variable Rate Particle Filter} \label{sec:vrpf}

The variable rate particle filter (VRPF) is described in \cite{Godsill2007,Godsill2007a}. The objective of the algorithm is to sequentially estimate the posterior distribution of the changepoint sequence, $p(\theta_{n}| y_{1:n})$, at each time $t_n$. This distribution may be expanded using Bayes' rule.
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(\theta_{n}|y_{1:n})} \nonumber \\
 & \propto & p(y_n|\theta_{n}, y_{1:n-1}) p(\theta_{n \setminus n-1}|\theta_{n-1}) p(\theta_{n-1}|y_{1:n-1}) \label{eq:vrpf_target}
\end{IEEEeqnarray}

The transition term, $p(\theta_{n \setminus n-1} | \theta_{n-1})$, can be constructed in a similar manner to (\ref{eq:cp_sequence_prior}) \cite{Jacobsen2006}, as a product of density terms for each new changepoint and a survivor function term. In the particular (but common) case that no new changepoints occur within the interval, the density consists of a probability mass on the empty set, with weight $P(\tau_{K_n+1} > t_n | \tau_{K_n+1} > t_{n-1}, \tau_{K_n} )$.
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1})} \nonumber \\
    & = & \begin{cases} S(\tau_{K_n}, t_n) \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1}, \tau_j>t_{n-1}) & K_n > K_{n-1} \\
                        S(\tau_{K_n}, t_n) / S(\tau_{K_n}, t_{n-1}) & K_n = K_{n-1} \end{cases} \IEEEeqnarraynumspace \label{eq:cp_sequence_trandens}
\end{IEEEeqnarray}

For all but the first changepoint in the interval, the density is given by the prior model of (\ref{eq:cp_model}). For the first changepoint, indexed by $k=K_{n-1}+1$, we must account for the fact that a changepoint cannot occur before $t_{n-1}$,
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(\tau_{k}, u_{k}| \tau_{k-1}, u_{k-1}, \tau_{k}>t_{n-1})} \nonumber \\
  & = & \frac{1}{S(\tau_{k-1}, t_{n-1})} \begin{cases} p(\tau_{k}, u_{k}| \tau_{k-1}, u_{k-1}) & \tau_{k} > t_{n-1} \\ 0 & \tau_{k} < t_{n-1} \end{cases}  \IEEEeqnarraynumspace \label{eq:cp_cond_model}   .
\end{IEEEeqnarray}

Practically, because changepoints will be relatively rare events, it is not likely that more than one new changepoint will occur between $t_{n-1}$ and $t_n$.

The target distribution of (\ref{eq:vrpf_target}) cannot be calculated analytically, but may be approximated numerically. A particle filter is an algorithm for approximating a probability distribution using a set of weighted samples (or ``particles'') drawn from that distribution. In this case, each particle will be a set of changepoint times and parameters.
%
\begin{equation}
 \hat{p}(\theta_{n}|y_{1:n}) = \sum_j w_n^{(j)} \delta_{\theta_{n}^{(j)}}(\theta_{n})
\end{equation}

where $\delta_x(X)$ is a dirac probability mass at $X=x$.

The particle filter works recursively. At the $n$th step, a particle, $\theta_{n-1}^{(i)}$, is first resampled from those approximating the filtering distribution at the $(n-1)$th step, using an appropriately chosen set of proposal weights.
%
\begin{equation}
 q(\theta_{n-1}) = \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{(j)}}(\theta_{n-1})
\end{equation}

The choice of weights determines the type of resampling used. The simplest choice, $v_{n-1}^{(j)} = 1/N_F$ (where $N_F$ is the number of filter particles) may be achieved by simply omitting this step all together and using the particles of $\hat{p}(\theta_{n-1}|y_{1:n-1})$. This, however, leads to degeneracy of the particle weights over time. Conventional resampling is achieved by using $v_{n-1}^{(j)} = w_{n-1}^{(j)}$. Any other choice results in an auxiliary particle filter \cite{Pitt1999}. For further discussion of resampling, see \cite{Cappe2007,Doucet2009,Douc2005}.

Next, an extension to the changepoint sequence, $\theta_{n \setminus n-1}^{(i)}$, is proposed from an importance distribution, $q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n)$, and concatenated to $\theta_{n-1}$ to create an estimate of $\theta_n$. Finally, the particle is weighted according to the ratio of the target and proposal densities.
%
\begin{IEEEeqnarray}{rCl}
w_n^{(i)} & = & \frac{ p(\theta_{n}^{(i)}|y_{1:n}) }{ q(\theta_{n}) } \nonumber \\
    & \propto & \frac{ p(y_n|\theta_n, y_{1:n-1}) p(\theta_{n \setminus n-1}|\theta_{n-1}) p(\theta_{n-1}|y_{1:n-1}) }{ q(\theta_{n-1}) q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n) } \nonumber \\
    & =       & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times \frac{ p(y_n|\theta_n, y_{1:n-1}) p(\theta_{n \setminus n-1}|\theta_{n-1}) }{ q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n) } \label{eq:vrpf_weights}
\end{IEEEeqnarray}

The normalisation may be enforced by scaling the weights so that they sum to $1$.

For the most basic ``bootstrap'' \cite{Gordon1993} form of the VRPF, $\theta_{n \setminus n-1}$ may be proposed from the prior transition density (\ref{eq:cp_sequence_trandens}). This can be achieved by sampling new changepoints sequentially from the transition model (\ref{eq:cp_model}) (apart from the first which is sampled from (\ref{eq:cp_cond_model})) until one falls after the current time, $t_n$. This final future changepoint is discarded. (This process can be thought of as sampling the entire future changepoint sequence from $t_{n-1}$ onwards, and then marginalising those which fall after $t_n$.) The bootstrap proposal leads to the usual simplification of the weight formula.
%
\begin{IEEEeqnarray}{rCl}
w_n^{(i)} & = & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times p(y_n|\hat{x}_n) \label{eq:bootstrap_vrpf_weights}
\end{IEEEeqnarray}

The choice of proposal weights, $\{v_{n-1}^{(i)}\}$, requires particular attention in the design of VRPFs. In some models a changepoint may not have an immediate effect on the observations, especially if a jump occurs in some quantity which is only observed via its integral, e.g. if there is a jump in the acceleration of a moving object, yet only the position is measured, the change will not be apparent until several more observations have been made. In the meantime, particles which contain a changepoint at the correct time may all have been removed by the resampling process. To avoid this loss of good particles, proposal weights should be chosen which preserve a significant number of low-weight particles. One scheme which has been found to work well is described in \cite{Godsill2007}, in which proposal weights are given by:
%
\begin{IEEEeqnarray}{rCl}
v_{n-1}^{(i)} & \propto & \max ( 1, N_F w_{n-1}^{(i)} )
\end{IEEEeqnarray}

It only remains to consider the likelihood term required for evaluation of the importance weights, $p(y_n|\theta_n, y_{1:n-1})$. The form of this term depends on the particular model under consideration. In the following sections, the likelihood expressions for the conditionally linear-Gaussian and deterministic cases are considered.

\subsection{Conditionally Linear-Gaussian Likelihoods} \label{sec:rb-vrpf}

For a conditionally linear-Gaussian model, the required likelihood term, $p(y_n|\theta_n, y_{1:n-1})$ is the predictive distribution estimated by the Kalman filter. Conveniently, the Kalman filter also provides us with an estimate of the current state given the changepoint sequence and all the preceding observations. It follows from the the Gaussian dynamics and prior that these distributions are all Gaussian as well, \cite{Anderson1979}, 
%
\begin{IEEEeqnarray}{rCl}
 p(x_n|\theta_{n}, y_{1:n-1}) & = & \mathcal{N}(x_n|m_n^-, P_n^-) \\
 p(x_n|\theta_{n}, y_{1:n}) & = & \mathcal{N}(x_n|m_n, P_n) \\
 p(y_n|\theta_{n}, y_{1:n-1}) & = & \mathcal{N}(y_n|\mu_n, S_n)     ,
\end{IEEEeqnarray}

with means and variances given by the following standard recursions (dependence on $\theta_{n}$ suppressed for clarity),
%
\begin{IEEEeqnarray}{rCl}
 m_n^- & = & A_n m_{n-1} \label{eq:kf_predict_start} \\
 P_n^- & = & A_n P_{n-1} A_n^T + Q_n \\
 \mu_n & = & C_n m_n^- \\
 S_n   & = & C_n P_n^- C_n^T + R_n \label{eq:kf_predict_stop} \\
 K_n   & = & P_n^- C_n^T S_n^{-1} \label{eq:kf_update_start}\\
 m_n   & = & m_n^- + K_n (y_n - \mu_n) \\
 P_n   & = & P_n^- - K_n S_n K_n^T \label{eq:kf_update_stop}    .
\end{IEEEeqnarray}

This completes the requirements for the particle filter, resulting in the final algorithm shown in Fig.~\ref{alg:RBVRPF}. The concept of estimating the posterior for just the changepoint sequence with a particle filter rather than the joint posterior over changepoints and linear states is an example of Rao-Blackwellisation (see, e.g. \cite{Casella1996,Doucet2000}), so this algorithm is known as the Rao-Blackwellised variable rate particle filter (RB-VRPF) \cite{Godsill2007a,Christensen2012}.

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}$, where $\tau_0^{(i)}=0$ and $\{u_0^{(i)}\} \sim p(u_0)$.
\STATE For each $i$, initialise particle sufficient statistics, $m_0^{(i)}$ and $P_0^{(i)}$ with prior values.
\FOR{$n=1 \dots N$}
  \FOR{$i=1 \dots N_F$}
  	\STATE Sample a changepoint history $\theta_{n-1}^{(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{(j)}}(\theta_{n-1})$.
    \STATE Propose a sequence extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1} | \theta_{n-1}^{(i)})$.
    \STATE Add extension to changepoint sequence $\theta_n^{(i)} \gets \theta_{n-1}^{(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
    \STATE Predict state mean and covariance $\mu_n^{(i)}$ and $S_n^{(i)}$ using (\ref{eq:kf_predict_start})~to~(\ref{eq:kf_predict_stop}).
    \STATE Calculate weight $w_n^{(i)}$ using (\ref{eq:vrpf_weights}).
    \STATE Update state mean and covariance $m_n^{(i)}$ and $P_n^{(i)}$ using (\ref{eq:kf_update_start})~to~(\ref{eq:kf_update_stop}).
  \ENDFOR
  \STATE Scale weights such that $\sum_i w_n^{(i)}=1$.
\ENDFOR
\end{algorithmic}
}}
\label{alg:RBVRPF}
\caption{Rao-Blackwellised variable rate particle filter}
\end{figure}

\subsection{Conditionally Deterministic Likelihoods} \label{sec:pd-vrpf}

When a conditionally deterministic model is used, the state at observation time $t_n$ is specified by the changepoint sequence $\theta_n$ (plus the initial state, $x_0$), using (\ref{eq:disc_time_state_diff_eq}). Thus, the required likelihood term is simply given by,
%
\begin{IEEEeqnarray}{rCl}
 p(y_n|\theta_{n}, y_{1:n-1}) & = & p(y_n|\hat{x}_n)     .
\end{IEEEeqnarray}

This leads to a particle filter for variable rate models with conditionally deterministic dynamics, summarised in Fig.~\ref{alg:VRPF}. The resulting trajectories $x(t)^{(i)}$ for each particle are realisations from a piecewise-deterministic process, so this algorithm is named the piecewise-deterministic variable rate particle filter (PD-VRPF).

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}$, where $\tau_0^{(i)}=0$ and $\{u_0^{(i)}\} \sim p(u_0)$.
\FOR{$n=1 \dots N$}
  \FOR{$i=1 \dots N_F$}
  	\STATE Sample a history $\theta_{n-1}^{(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{(j)}}(\theta_{n-1})$.
    \STATE Propose a sequence extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_{n})$.
    \STATE Add extension to changepoint sequence $\theta_n^{(i)} \gets \theta_{n-1}^{(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
    \STATE Calculate state $\hat{x}_n$ using (\ref{eq:disc_time_state_diff_eq}).
    \STATE Calculate particle weight using (\ref{eq:vrpf_weights}).
  \ENDFOR
  \STATE Scale weights such that $\sum_i w_n^{(i)}=1$.
\ENDFOR
\end{algorithmic}
}}
\label{alg:VRPF}
\caption{Piecewise deterministic variable rate particle filter}
\end{figure}

\subsection{Improving the Variable Rate Particle Filter}

The bootstrap versions of the VRPFs described in sections~\ref{sec:rb-vrpf} and~\ref{sec:pd-vrpf} may perform poorly if changepoints are not obvious until significantly after they occur. For example, in a tracking example, if a jump occurs in the acceleration, but only the position is observed, then this change may not be obvious until a number of observations have arrived. In this case, the estimation may be improved by the introduction of resample-move (RM) steps \cite{Gilks2001}. In an RM scheme, optional Metropolis-Hastings (MH) moves are conducted to alter the particle states after the importance sampling has taken place. For variable rate models, any one of the previous changepoints, $\tau_k$, or associated parameters, $u_k$, could be adjusted. Because more observations are available than when the changepoint was first proposed, it may be possible to construct more informed proposals and so move the changepoints towards regions with higher posterior probability. It is even possible to retrospectively add or remove changepoints, using reversible jump MH moves \cite{Green1995}. Variable rate particle filters using RM with piecewise deterministic models are described in \cite{Whiteley2011,Gilholm2008}.

Rather than conducting the IS and MH steps separately, it is possible to combine them using the framework of SMC samplers \cite{DelMoral2006}. This was suggested in \cite{Whiteley2011}, again for piecewise deterministic dynamics, but the extension to conditionally linear-Gaussian models is straightforward.

Filtering schemes which alter past changepoints -- whether using an SMC sampler or RM -- are computationally expensive, because many likelihood calculations must be conducted, for each observation from the time of the change onwards. In some cases, it may be simpler to just use a bootstrap filter with more particles.

\section{Variable Rate Particle Smoothing} \label{sec:vrps}

A filter conducts inference sequentially as new observations are introduced. The purpose of a smoother is to produce a second estimate once all the observations have been made, using future values to improve upon the filter performance. Estimating changepoints online is a challenging task because the presence of a change may not be obvious until after it has happened. Thus, it is anticipated that a smoothing algorithm will provide significantly improved performance at changepoint estimation.

The target distribution for a variable rate smoothing algorithm is the posterior distribution over the entire changepoint sequence, $p(\theta|y_{1:N})$. A particle approximation to this distribution is generated by the final step of the VRPF. However, in the same manner as the fixed rate filter-smoother of \cite{Kitagawa1996}, this approximation is likely to lack path-space diversity -- because of the necessary resampling step in the filtering algorithm, the particles all share the same set of changepoints before a particular time, with variation only appearing for changepoints closer to $T$. For a good characterisation of the smoothing distribution, it is necessary to rejuvenate the set of particles. This is achieved with a backward pass through the observations in a similar manner to the forward-backward method described in \cite{Godsill2004}.

The target distribution may be factorised as follows,
%
\begin{IEEEeqnarray}{rCl}
 p(\theta|y_{1:N}) = p(\theta_{n}^{+}|y_{1:N}) p(\theta_{n}|\theta_{n}^{+}, y_{1:N})     .
\end{IEEEeqnarray}

This suggests a sequential strategy for smoothing. A particle drawn from $p(\theta_{n}^{+}|y_{1:N})$ may be extended backwards by sampling from the backwards conditional distribution, $p(\theta_{n}|\theta_{n}^{+}, y_{1:N})$ and concatenating the two partial sequences. The new particle will be a sample from the target distribution, but the resulting approximation will still suffer from low path-space diversity before time $t_{n-1}$. Therefore, $\theta_{n-1}$ is marginalised by simply discarding the changepoints which occur before this time. This yields in a particle drawn from $p(\theta_{n-1}^{+}|y_{1:N})$. The procedure then continues recursively. After a complete backwards pass through the observations, a single particle from the target distribution will have been generated. The process is then repeated until sufficient samples are obtained.

At step $n$ for a given particle, the future changepoint sequence may be considered to be fixed, and is denoted $\tilde{\theta}_n^+$. It remains to devise a method for drawing samples from the backwards conditional distribution, $p(\theta_{n}|\theta_{n}^{+}, y_{1:N})$. Such methods are discussed in the following sections for the two classes of models.

\subsection{Conditionally Linear-Gaussian Smoothing} \label{sec:rb-vrps}

When the state dynamics are linear-Gaussian conditional on the changepoint sequence, it is possible split the dependence on the observations into past and future by introducing the current state as an additional variable. This trick was used in \cite{Sarkka2012} in the derivation of the Rao-Blackwellised particle filter for fixed rate conditionally linear-Gaussian models.
%
\begin{IEEEeqnarray}{rCl}
 \IEEEeqnarraymulticol{3}{l}{p(\theta_{n}|\tilde{\theta}_{n}^+, y_{1:N})} \nonumber \\
\qquad & \propto & p(\theta_{n}, \tilde{\theta}_{n}^+| y_{1:N}) \nonumber  \\
       & =       & \int p(x_n, \theta_{n}, \tilde{\theta}_{n}^+| y_{1:N}) dx_n \nonumber  \\
       & \propto & \int p(y_{n+1:N}|x_n, \theta_{n}, \tilde{\theta}_{n}^+, y_{1:n}) p(x_n, \theta_{n}, \tilde{\theta}_{n}^+| y_{1:n}) dx_n \nonumber \\
       & = & \int p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+) p(x_n|\theta_{n}, y_{1:n}) dx_n \nonumber \\
       &   & \times p(\tilde{\theta}_{n}^+|\theta_{n}) p(\theta_{n}|y_{1:n})
\end{IEEEeqnarray}

Finally, the particle approximation is substituted for the filtering distribution.
%
\begin{IEEEeqnarray}{rCl}
\hat{p}(\theta_{n}|\tilde{\theta}_{n}^+, y_{1:N}) & = & \sum_i \tilde{w}_{n}^{(i)} \delta_{\theta_{n}^{(i)}}(\theta_{n})  \label{eq:rb-vrps_back_cond}
\end{IEEEeqnarray}
